Toward theories of deliberated preferences

### 2020-10-26 simple.tex
Review improved

It is by no means my intent to claim that descriptive approaches never face validation difficulties. Experimental tests to arbitrate between theories may not be feasible in practice, descriptions are not always solely about the future, falsificationist approaches in turn raise epistemological difficulties, and so on. Correspondingly, the possible confrontation to empirical data of the normative and descriptive theories defined here will likely not suffice to close every discussion about validation. My less naïve claim, that should suffice to motivate the present endeavor, is that it is often a good thing for a theory to make empirically testable claims, from the point of view of the possibilities of solving disputes and validating theories.

As the plural indicates, this article defines a kind of theories, called theories of \ac{DJ}, not specific theories of \ac{DJ}. Specific theories of \ac{DJ} are tailored to a set of situations and yield claims relevant to those situations. For example, a specific theory of \ac{DJ} may be tailored to situations of choice between lotteries involving money. This article adopts a methodological point of view. It proposes a definition of what theories of \ac{DJ} should look like, and proves claims that hold for any specific theory of \ac{DJ}, showing that such theories are indeed normative and descriptive.

\Cref{sec:approaches} defines more precisely what is meant here with normative and descriptive theories. These definitions permit to readily see why the apparent impossibility of a claim being both normative and descriptive is an illusion, thereby enabling a solution to what could be called the “normative against descriptive” gap.

\section{Species of theories}
\label{sec:approaches}
\subsection{Overview}
For this article, a \emph{normative decision theory} is a theory that, when applied in the situations it self-describes as applicable to, claims to yield valid prescriptions. I consider a prescription in the wide sense: recommending against some behavior, or recommending conditionally, also counts as prescribing.

@@

The descriptive theories proposed in decision theory, in the sense defined here, are exclusively shallow descriptive theories, to the best of my knowledge.
%This need not be so. Indeed, this article introduces descriptive theories about reflective behavior.
This observation permits to understand what could be called the “normative versus descriptive gap”, implicit in the decision theoretic literature: because normative theories aim to prescribe, they are required to not yield recommendations that precisely, systematically match shallow behavior. Given the numerous results in experimental psychology showing that shallow behavior is undefensible in many situations, a normative theory whose prescriptions would systematically match shallow behavior would be suspicious, to say the least. It would also be of low practical interest.
It is an immediate consequence of this minimal requirement that a normative theory cannot also be a shallow descriptive theory. 
In other words, as \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow behavior is not sufficient to inform normative economics.
The “normative versus descriptive gap” comes from an implicit identification of descriptive theories with shallow descriptive theories.

@@

\subsection{Existing approaches for modeling reflective preferences}
The review of existing approaches aiming at obtaining a model of the reflective preferences of an individual distinguishes two kinds of approaches. \Cref{sec:purification} mentions formal techniques, that can be applied by following an explicit procedure, but that do not consider the reaction of individuals to arguments; while \cref{sec:informal} reviews studies that involve informal discussions with individuals, which do consider arguments, but require involving experts. By contrast, this article presents theories that can be confronted to empirical data in a formal, systematic way, and that consider the reactions of individuals to arguments.

@@

\subsubsection{Preference purification}
\label{sec:purification}
An approach much discussed recently is the one of preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}.
Having collected shallow preference data (from the field or from interviews), purifying preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. 
For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet[Section VII]{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information.%; \citet{rubinstein_eliciting_2011} study the possible “rational” preference relations that are coherent with the observed data when assuming some distortion process relating (shallow, observed) and (reflective, unobserved) behavior. 
\commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}
\commentOC{Je ne vois pas comment faire ça sans en faire une digression énorme. Il faudrait introduire des notations, des définitions, … Je vois à peu près comment les présenter sous un cadre unifié et les comparer axiomatiquement aux théories du DJ, mais cela devrait faire l’objet d’un autre article, il me semble.}

@@

\subsubsection{Informal deliberation}
\label{sec:informal}
Another path to obtain reflective preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of them \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011, chappell_deliberative_2012}, aiming to obtain well-informed preferences. 
The term as used here also includes the informal use of debiasing techniques: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). 
\commentYM{il me semble nécessaire ici de mentionner aussi tous les dispositifs participatifs et de concertation employés en routine dans les prises de décisions relatives aux politiques publiques et leur mise en oeuvre}
\commentOC{Je m’interroge. Ça me semble de pus en plus éloigné du scope, car je m’intéresse aux préférences d’un individu donné, hors questions d’agrégation.}

A major problem of these techniques is that they are not easily reproducible \commentYM{il faudrait expliquer en quel sens}: the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals.

This problem is compounded by the phenomenon of arbitrary coherence, exhibited by psychologists \citep{ariely_predictably_2010}.  This term denotes situations where individuals form an apparently coherent system of preference that depends on the path taken to elicit it. 
\commentYM{phrase beaucoup trop longue pour être compréhensible, qui fait par ailleurs référence au concept de rationalité sans le définir, ce qui est acceptable en début d'intro, mais me semble problématique à ce stade dans un papier qui cherche à éclairer les débats flous par des clarifications formelles et conceptuelles}.

@@

\commentOCf{C’est discutable si c’est un autre problème ou un peu le même problème ou carrément le même problème vu autrement. Du point de vue logique, toutes les propositions fausses sont équivalentes. En l’absence d’une analyse comparative approfondie de toutes ces approches et de leurs propriétés, je ne vois pas à ce stade comment faire beaucoup plus clair. Mais je serais heureux d’en discuter si tu vois une manière de patcher le problème raisonnablement facilement.}

@@

\subsection{Some other related approaches}
\commentYM{related to what? to your approach or to the one explored above?}
The current proposal pursues a path started with \citet{cailloux_formal_2019}, an article that defines the notion of \ac{DJ} and models that enable interactions with individuals whose \acp{DJ} are to be captured. That article however requires as fundamental observable a “trump” relation, indicating whether an individual thinks that an argument is against another argument; and a “support” relation, indicating whether an individual thinks that an argument supports a proposition.\commentYM{je ne pense pas qu'on puisse tenir pour acquis que le lectorat connait par coeur l'article delibenv. Il faut du coup donner plus d'explication. Par ailleurs, il est impératif que tu expliques pourquoi tu considères qu'il n'est pas raisonnable de considérer que trump et support sont observables} The current proposal defines every required concepts, such as the notion of arguments attacking other arguments, on the basis of a single observable that is compatible with the revealed preference approach priviledged in economic theory \commentYM{je ne pense pas qu'on puisse dire que l'approche RP est privilégiée en économie. Elle est aussi très critiquée. Si tu la considère particulièrement solide, il faut expliquer pourquoi}, in the sense that it can be defined by reference to what individuals effectively choose under various circumstances. The current proposal also defines explicitly (in the form of an axiom \commentYM{il n'y QU'UN axiome au final ?}) sufficient requirements to grant normative power to observations, an aspect which was previously left unclear. Finally, the current proposal links \ac{DJ} and deliberated preferences. \commentYM{indispensable de dire explicitement quelle est la différence entre les deux dès maintenant}

@@

\subsection{This approach}\commentYM{The proposed approach}
The proposal of this article \commentYM{the proposal developed/elaborated/presented in this article}to capture reflective preferences while avoiding the just described problems is to define normative theories on the basis of a low demanding axiom complemented with principles of reasonings; \commentYM{je devine que cette phrase est importante mais elle ne se comprend pas. Sans le ";" ici elle se comprendrait. Mais est-ce que c'est comme ça qu'il faut la comprendre ? ". This" à la place de "; that" donne un autre sense : lequel est le bon ?}that allow decision makers to “reconstruct” \commentYM{pourquoi ce terme est-il mis entre guillemets ? généralement on fait cela quand on veut souligner que le terme doit être compris en un sens différent d'un sens habituel proéminent. Mais il me semble pas que "reconstruct" ait un sens habituel proéminent dans ce contexte. Donc je bloque et je me dis "j'ai loupé un épisode"}the reasoning proposed by the principles so as to be able to check that they apply to their situation; therefore permitting to falsify the theories using observable data representing the rejection by the decision makers of these principles \commentYM{phrase à ralonge, vraiment pas idiomatique en anglais, et à peu près impossible à suivre (du moins pour les lecteurs de ma catégorie)}. Such theories are therefore both normative and empirical: normative, in the sense of claiming to capture reflective preferences, which provide a better basis for recommendations or policy making than shallow preferences; and empirical, in the sense of the principles of reasoning being exposed to empirical tests.

The proposed approach is based on a concept of preference that I call deliberated preference, a formal refinement of the vague notion of reflective preferences. Those are the preferences of an individual that are stable facing counter-arguments. The arguments are presented to the decision maker, and her reaction is observed, for example, in the form of a choice between some alternatives. This is the basic observable of interest to the theories here defined.

It is not an ambition of this article to actually propose concrete theories of deliberated preferences, for this is not achievable in a single article; and the theories as defined here admittedly still suffer important weaknesses. \commentYM{ce qui précède n'est pas compréhensible si t n'explicites pas ce que veut dire "theory" dans cet article. Par ailleurs, qu'est-ce qu'une theorie CONCRETE ? A quoi cela s'oppose-t-il ? Le lecteur se demande aussi en quoi cela consiste de "définir" une théorie (ce que tu sous-entends que tu vas faire) sans la "proposer" (ce que tu dis dépasser tes ambitions). Enfin, de quelles "weaknesses" parles-tu ?}The ambition of this article is accordingly limited to present a novel approach of capturing reflective preferences in simple enough settings.\commentYM{quel sens faut-il donner à "setting" ici ? dans certains cas, on peut employer ce genre de terme de manière vague sans que ça choque. Ici je pense que ce n'est pas possible, car le lecteur ne comprend pas où tu vas, ni ce que tu vas faire, ni pourquoi c'est compliqué, ni quelle est la complexité dont tu parles, ni ce qui peut faire qu'un setting est "simple enough" ou complexe}
%This requires to define precisely what such theories should aim to capture; and how they should be compared, which is the sole ambition of this article. 
Some challenges that must be overcomed \commentYM{overcome}in order to build concrete theories applicable in realistic settings will be discussed in conclusion.\commentYM{tu pourrais mettre ce genre de phrase à la fin d'une annonce de plan, mais là comme ça, elle ne sert à rien}

One long-term goal of this research program is to give individuals the means to reject theories that they find unconvincing after due consideration of all arguments; and accordingly, give incentives to decision theorists to build defenses of these theories able to convince individuals whose preferences are modeled of the soundness of these theories. \commentYM{phrase trop longue}This might match \possessivecite{hausman_debate_2010} suggestion that we should use rational persuasion rather than nudge individuals, arguing that it would promote autonomy.
It also follows \possessivecite[p.\ 72]{guala_logic_2000} suggestion (discussing the Allais paradox) about obtaining relevant data to compare normative behavioral theories: “Before asking whether agents want to revise their choices or not, \emph{both} rival theories should be proposed” (emphasis in the original text).

@@

\section{Informal presentation}
\commentYM{Le lecteur n'a aucune idée de ce qu'il doit s'attendre à trouver dans cette section, et de la mesure dans laquelle ça va permettre d'avancer dans la résolution de la problématique : lecteur perdu = pas bien}
Although the final aim is about the concept of deliberated preference, which permits to compare alternatives (options in a decision problem), this article will mostly discuss \acp{DJ}. That is because, once the concept of \ac{DJ} is clearly established, it is a simple matter to define deliberated preferences, and will be done in a later section, by viewing a preference instance as a judgment about a pair of alternatives.

For a clear separation of concern, this article defines first normative theories, then empirical theories for these normative theories \commentYM{je ne comprends pas ce "for" ; je ne comprends pas pourquoi tu utilises une autre opposition que la classique normative vs. empirical. Par ailleurs, je ne comprends pas DE QUELLES théories tu parles. Es-tu en train de parler des théories exitantes, ou des théories dont le développement serait une partie de la réalisation du projet du présent papier ? Comme déjà dit plus haut, en fait je ne sais pas ce que tu appelles une "theory"}. A normative theory makes normative claims only, which have to be accepted or rejected on the basis of the apparent soundness of the axiom on which it rests \commentYM{par qui ? dans le cadre de l'analyse ou en amont ? et pourquoi sur la seule base de la soundness de l'axiome ? pourquoi pas sur la base de ses implications dans des cas concrets ?}. An empirical theory for a given normative theory, by contrast, makes empirical claims, hence, can be confronted to observations \commentYM{quelle différence avec les implications empiriques d'une théorie normative ?}. What I call here an empirical theory, for simplicity, is not however purely empirical, but can also be considered normative, for two reasons. First, its conclusions inevitably rest on the normative validity of the normative theory for which it is built. Second, it claims to capture preferences that provide an appropriate basis for recommendation. Empirical theories of \ac{DJ} could thus be called empirical normative theories of \ac{DJ}.\commentYM{est-ce la même chose que ce qui est déjà appelé à peu près comme ça dans la littérature ? Je parle par exemple de toutes les théories empiriques de la justice}

>>>>

It is by no means my intent to claim that descriptive approaches never face validation difficulties. Experimental tests to arbitrate between theories may not be feasible in practice, descriptions are not always solely about the future, falsificationist approaches in turn raise epistemological difficulties, and so on. Correspondingly, the possible confrontation to empirical data of the theories defined here will likely not suffice to close every discussion about validation. My less naïve claim, that should suffice to motivate the present endeavor, is that it is often a good thing for a theory to make empirically testable claims, from the point of view of the possibilities of solving disputes and validating theories.

As the plural indicates, this article defines a kind of theories, called theories of \ac{DJ}, not a specific theory of \ac{DJ}. Specific theories of \ac{DJ} are tailored to a set of situations and yield claims relevant to those situations. For example, a specific theory of \ac{DJ} may be tailored to situations of choice between lotteries involving money. This article adopts a more general point of view. It proposes a definition of what theories of \ac{DJ} should look like, and proves claims that hold for any specific theory of \ac{DJ}, showing that such theories are indeed normative and descriptive.

\Cref{sec:approaches} defines more precisely what is meant here with normative and descriptive theories. These definitions permit to readily see why the apparent impossibility of a claim being both normative and descriptive is an illusion, thereby enabling a solution to what could be called the “normative against descriptive” gap. It also reviews related works in the literature, and briefly sketches the current proposal.
\Cref{sec:normative,sec:descriptive} define the theories of \ac{DJ} and prove that these theories are normative and descriptive in the sense defined in \cref{sec:approaches}. \Cref{sec:preferences} define theories of \ac{DP} on the basis of theories of \ac{DJ}, and \cref{sec:conclusion} concludes.

\section{Eliciting reflective preferences}
\label{sec:approaches}
\subsection{Species of theories}
For this article, a \emph{normative decision theory} is a theory that, when applied in the situations it self-describes as applicable to, claims to yield valid prescriptions. I consider a prescription in the wide sense: recommending against some behavior, or recommending conditionally, also counts as prescribing.

@@

The descriptive theories proposed in decision theory, in the sense defined here, are exclusively shallow descriptive theories, to the best of my knowledge.
%This need not be so. Indeed, this article introduces descriptive theories about reflective behavior.
This observation permits to understand what could be called the “normative versus descriptive gap”, referring to the belief, implicit in the decision theoretic literature, that a theory cannot be both normative and descriptive. Because normative theories aim to prescribe, they are required to not yield recommendations that precisely, systematically match shallow behavior. Given the numerous results in experimental psychology showing that shallow behavior is undefensible in many situations, a normative theory whose prescriptions would systematically match shallow behavior would be suspicious, to say the least. It would also be of low practical interest.
It is an immediate consequence of this minimal requirement that a normative theory cannot also be a shallow descriptive theory. 
In other words, as \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow behavior is not sufficient to inform normative economics.
The “normative versus descriptive gap” comes from an implicit identification of descriptive theories with shallow descriptive theories.

@@

The review of existing approaches aiming at obtaining a model of the reflective preferences of an individual distinguishes two kinds of approaches. \Cref{sec:purification} mentions formal techniques, that can be applied by following an explicit procedure, but that do not consider the reaction of individuals to arguments; while \cref{sec:informal} reviews studies that involve informal discussions with individuals, which do consider arguments, but require involving experts. By contrast, this article presents theories that can be confronted to empirical data in a formal, systematic way, and that consider the reactions of individuals to arguments.

@@

\subsection{Preference purification}
\label{sec:purification}
An approach much discussed recently is the one of preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}.
Having collected shallow preference data (from the field or from interviews), purifying preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. 
For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet[Section VII]{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information.%; \citet{rubinstein_eliciting_2011} study the possible “rational” preference relations that are coherent with the observed data when assuming some distortion process relating (shallow, observed) and (reflective, unobserved) behavior. 
\commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}
\commentOC{Je ne vois pas comment faire ça sans en faire une digression énorme. Il faudrait introduire des notations, des définitions, … Je vois à peu près comment les présenter sous un cadre unifié et les comparer axiomatiquement aux théories du DJ, mais cela devrait faire l’objet d’un autre article, il me semble.}

@@

\subsection{Informal deliberation}
\label{sec:informal}
Another path to obtain reflective preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of them \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011, chappell_deliberative_2012}, aiming to obtain well-informed preferences. 
The term as used here also includes the informal use of debiasing techniques: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). 
\commentYM{il me semble nécessaire ici de mentionner aussi tous les dispositifs participatifs et de concertation employés en routine dans les prises de décisions relatives aux politiques publiques et leur mise en oeuvre}
\commentOC{Je m’interroge. Ça me semble de pus en plus éloigné du scope, car je m’intéresse aux préférences d’un individu donné, hors questions d’agrégation.}

A major problem of these techniques is that they are not easily reproducible \commentYM{il faudrait expliquer en quel sens}, because they require informal interactions with experts. Furthermore, the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals.

This problem is compounded by the phenomenon of path dependency, or arbitrary coherence, exhibited by psychologists \citep{ariely_predictably_2010}.  This term denotes situations where individuals form an apparently coherent system of preference that depends on the path taken to elicit it. 
\commentYM{phrase beaucoup trop longue pour être compréhensible, qui fait par ailleurs référence au concept de rationalité sans le définir, ce qui est acceptable en début d'intro, mais me semble problématique à ce stade dans un papier qui cherche à éclairer les débats flous par des clarifications formelles et conceptuelles}.

@@

\commentOCf{C’est discutable si c’est un autre problème ou un peu le même problème ou carrément le même problème vu autrement. Du point de vue logique, toutes les propositions fausses sont équivalentes. En l’absence d’une analyse comparative approfondie de toutes ces approches et de leurs propriétés, je ne vois pas à ce stade comment faire beaucoup plus clair. Mais je serais heureux d’en discuter si tu vois une manière de corriger le problème raisonnablement facilement.}

@@

\subsection{Previous proposal to study \aclp{DJ}}
The current proposal pursues a path started with \citet{cailloux_formal_2019}, an article that defines the notion of \ac{DJ} and describe how \acp{DJ} can be observed.

That previous proposal postulates that an individual $i$, whose reflective preferences are being modeled, can adopt multiple perspectives. The proposal postulates that the modeler can observe two relations by interacting with $i$, called support and trump. They constitute the argumentative disposition of an individual $i$ and are defined as follows.

An argument $\ar$ supports a proposition $\prop$ iff $i$ considers that $\ar$ is in favor of $\prop$. This notion is conditional, in the sense that it does not say anything about the validity of $\ar$; only that if $\ar$ holds, then $i$ should endorse $\prop$.

An argument $\ar_1$ trumps another argument $\ar_2$ iff there is at least one perspective within which $i$ considers that $\ar_2$ turns $\ar_1$ into an ineffective argument. (This notion is formally decomposed into two relations, “trump in some perspective” and “does not trump in some perspective”, but these details do not need to be discussed here.)

That proposal then defines a number of notions, including the central notion of the \ac{DJ} of $i$, on the basis of her argumentative disposition. An individual’s \ac{DJ} is the set of propositions that resist counter-arguments, in some technical sense defined on the basis of these relations.

The current article improves on several aspects over this previous proposal.

First, the previous proposal requires the individual to declaratively take position on notions that may be unclear to him, as they include some possibly counterfactual situations, and as the concept of turning an argument into an uneffective argument may not be understood in the same way by everybody. 
In order for the conclusions about $i$’s \ac{DJ} to hold, we must be sure that $i$ understands the semantics of the relations on which it is based as we do. It is therefore desirable that these notions have unambiguous semantics. It is undoubtedly an important strength of \citet{samuelson_foundations_1983}, \citet{von_neumann_theory_2004} or \citet{savage_foundations_1972}’s theories that they use as primitive notions observable entities whose semantics is unambiguous. This is not to say that these proposals cannot be criticized. For example, the claim that the notion of preference can be disposed of entirely has been much criticized \citep{hausman_preference_2011, wong_foundations_2006}, with good reasons; and observations of shallow behavior, though having a clear semantics, are insufficient to deduce normative claims, as has been amply remarked and as argued in this very article. This is also not to say that declarations should never be used and that only choice acts should matter to any theory in economics.
Nonetheless, as \citet[p.\ 16]{von_neumann_theory_2004} aptly summarize, “It is clear that every measurement – or rather every claim of measurability – must ultimately be based on some immediate sensation, which possibly cannot and certainly need not be analyzed any futher”. A weakness of the previous proposal for studying \acp{DJ} is that the observable relations it uses may be considered inadequate primitive notions. This article improves on this aspect by postulating a non-ambiguous observable relation, that does not require $i$ to take position declaratively on relative qualities of arguments; but rather to choose (by declaration of preference or by act) among bundles, or to declare whether $i$ adheres to a given proposition in her current perspective. (In the latter case, the clarity of the semantics of the observable relation will depend on the clarity of the propositions.)

Another central aspect of the study of \acp{DJ} is that individuals may change perspective, thus, change opinion about whether an argument trumps another one. An important case of change of perspective is when the current perspective depends on which arguments $i$ has seen so far. The previous proposal studies these changes of perspectives through the trump relation, without modeling explicitly the arguments seen so far. This has the advantage of simplicity, but the price to pay is a lack of ability to reason explicitly about path dependency mechanisms: individuals may form stable opinions that differ according to, for example, the first argument they have seen. In such a case, although their opinion may stabilize (in the sense of resisting counter-arguments), it is still arbitrary. This article models paths (sequences of arguments) explicitly and prohibits such situations of arbitrariness to count as forming \acp{DJ}.

This article also clarifies the normative and descriptive status of the theories of \acp{DJ}. It introduces an explicit axiom which gives them their normative strength. This axiom is supplemented by claims, which are shown to be falsifiable by empirical data. The previous proposal defines conditions and discusses several possible interpretations of them, including empirical ones, but leaves open the question of how to test these conditions systematically using the observable relations.

\commentYM{la vision vNM est très réductrice, et ne peut pas être considérée, me semble-t-il, comme valable pour toutes les théories normatives, car l'idée qu'il doit y avoir des sensations, atomiques et immédiates relève de l'empirisme logique qui sous-tend leur approche, et cette épistémologie, ou ontologie à ce stade, est loin d'être partagée par toutes les approches à même de développer des théories normatives. Ta phrase qui suit risque de n'être pas beaucoup plus convaincantes pour les lecteurs, je le crains. Imaginons par exemple une théorie marxiste, elle s'appuiera sur des intérêts matériels, et se contre fout des réactions des individus aux arguments. On peut envisager par ailleurs des théories supra-individualistes, qui se foutent des réactions des individus. Mais bien entendu la pertinence de ce que je dis là dépend de ta réponse à ma première question dans l'intro du 2. Si ce dont tu parles ici, ce n'est pas des théories normatives existantes ou possibles, mais uniquement des théories normatives QUE TU ACCEPTES COMME ELEMENTS DE TA CONSTRUCTION, alors le problème est différent : il faut justifier TON adhérence à l'empirisme logique, ce qui n'est pas une mince affaire étant donné le discrédit qui lui est aujourd'hui attaché} 
\commentOCf{Je soupçonne que le problème était dû à une formulation maladroite de ma part (\og{}The fundamental element of a normative theory of deliberated preferences is the reaction of an individual, whose preferences are to be captured, to arguments\fg). Si ma reformulation ne résoud pas le problème, je suis intéressé qu’on en discute.}

\commentYM{je ne pense pas qu'on puisse tenir pour acquis que le lectorat connait par coeur l'article delibenv. Il faut du coup donner plus d'explication. Par ailleurs, il est impératif que tu expliques pourquoi tu considères qu'il n'est pas raisonnable de considérer que trump et support sont observables} 
\commentYM{je ne pense pas qu'on puisse dire que l'approche RP est privilégiée en économie. Elle est aussi très critiquée. Si tu la considère particulièrement solide, il faut expliquer pourquoi}
\commentYM{il n'y QU'UN axiome au final ?}

Finally, the current proposal links \ac{DJ} and deliberated preferences. \commentYM{indispensable de dire explicitement quelle est la différence entre les deux dès maintenant}

@@

\subsection{The proposed approach}
The proposal elaborated in this article defines theories that are both normative, and descriptive of reflective behavior.
The kind of behavior of interest here is choice behavior, although the proposal is more general.

In fact, the main object of study of this article is judgments about propositions. The main goal is accordingly to study reflective adherence to propositions.
This will then be applied to determining reflective choice behavior. Choice behavior is obtained when considering adherence to propositions of the form: “I ought to choose alternative $\alt$ rather than $\alt'$ if both are available”. The alternatives can be concrete objects such as consumable goods.

One of our main preoccupation will be to clarify, using formal and precise definitions, the intuitive notion of reflective adherence to propositions. This is done using the concept of \emph{deliberated} judgments (a normative concept) and of \emph{acceptable} propositions (a descriptive concept).

A theory of \ac{DJ} applies to a set of individuals and a set of propositions. The \ac{DJ} of an individual $i$ is the set of propositions $\prop$ such that it is legitimate to recommend to $i$ adherence to $\prop$. Someone’s \ac{DJ} is not directly observable.
However, we can observe the reactions of $i$ to arguments in favor and against the propositions.
A theory of \ac{DJ} assumes that it knows a set of arguments such that exploring the reactions to these arguments “sufficiently” permits to obtain knowledge about someone’s \ac{DJ}.
Thus, one important task dealt with in \cref{sec:normative} will be to define the notion of acceptability, that serves to capture formally this notion of “sufficient” exploration. Intuitively, a proposition is acceptable when some argument makes $i$ seem (according to the observations) to adhere to the proposition whatever the counter-arguments given to $i$, and whatever the sequence of arguments given to $i$ previously.

A theory of \ac{DJ} must use an axiom that relates the descriptive notion of acceptability to the normative notion of being deliberated. This axiom simply mandates that whatever is acceptable, is deliberated. \Cref{sec:normative} presents it formally.

This defines theories of \ac{DJ} that can be considered normative, as accepting the axiom is sufficient to determine unambiguously a set of legitimate prescriptions to be given to individuals. 
This is done by linking \acp{DJ} to acceptable propositions, a notion whose major interest is that it can be confronted to empirical observations. But at this stage, it is still unclear how to obtain knowledge of someone’s \ac{DJ}, and accordingly, such theories cannot still be considered as descriptive. In particular, they are not falsifiable, as they take no position on the reactions of individuals to arguments. \Cref{sec:descriptive} accordingly defines descriptive theories of acceptable propositions, assuming given a normative theory of \ac{DJ}.

Such descriptive theories can take multiple forms, and some of them would permit to see clearly that the theory is descriptive in the required sense, that is, that its claims can be falsified given observational data. 
But the task we face, for such theories to be useful, is more complex than this minimal requirement: we have to find forms of theories that are not hugely demanding on the part of the theorist. In particular, we do not want to oblige the theorist to predict the whole of the argumentative attitude of individuals if this is not required to obtain information about her \ac{DJ}. Thus, \cref{sec:descriptive} defines descriptive theories of \ac{DJ} that only claim to be able to answer counter-arguments, in a sense that will be made precise. The desire that its claim be falsifiable is also defined formally. More formally, the result that is obtained in that section, and that is the central result of this article, is that if a theory is convincing and normatively adequate, then it is valid. That a theory is convincing is (as also shown in the same section) the same as demanding that it resists all empirical falsification attempts. It is normatively adequate if it satisfies the normative axiom described here above informally (and formally in \cref{sec:normative}). And it is valid if the propositions that it claims are in someone’s \ac{DJ} are indeed in that person’s \ac{DJ}. This central result thus amounts to prove the main claim of this article, namely, that accepting a weak normative axiom, supplemented with empirical claims, is enough to capture someone’s reflective preferences.

A possible objection to the usefulness of this approach should perhaps be lifted before delving into the formal descriptions of this rough sketch.
It may be argued that because this proposition needs a normative axiom, which cannot be validated empirically, it does not fare better than the alternative propositions examined in the previous sections. \ac{SEU}, for example, also claim to obtain reflective preferences on the basis of normative axioms.
I grant that it is impossible to get completely rid of normative axioms, because prescriptions must be linked to descriptions, and this cannot be done in a complete empirical way.
However, it is good to lighten as much as possible the content of normative axioms, and defer as much content as possible to descriptive claims. The descriptive parts of theories can be confronted to observations, a natural way to lift disagreements that has often proved useful in the history of science (again, accepting this claim does not require to ignore that observations themselves rest upon theories). Arguably, the normative axiom proposed here is much weaker than those of a theory such as \ac{SEU}, in the sense of having less substance or of being compatible with more preference relations. (This claim can only be valid at an intuitive level: because \ac{SEU} does not use arguments, its axioms cannot be compared formally to mine.) This does not prevent theories of \ac{DJ} to propose substantive, non-trivial claims about reflective preferences. It simply moves the debate from the world of discussions between experts about the validity of the axioms to the world of observations of acceptability of arguments by the persons to which these theories claim to be applicable.
It is also worth noting that this axiom has a specific form that may permit decision makers to reconstruct the reasoning proposed by the theory so as to be able to check that they apply to their situation. Detailed exploration of this aspect is deferred to future work.

One long-term goal of this research program is to give individuals the means to reject theories that they find unconvincing after due consideration of all arguments; and accordingly, give incentives to decision theorists to build defenses of these theories able to convince individuals of the soundness of these theories. \commentYM{phrase trop longue}This might match \possessivecite{hausman_debate_2010} suggestion that we should use rational persuasion rather than nudge individuals, arguing that it would promote autonomy.
It also follows \possessivecite[p.\ 72]{guala_logic_2000} suggestion (discussing the Allais paradox) about obtaining relevant data to compare normative behavioral theories: “Before asking whether agents want to revise their choices or not, \emph{both} rival theories should be proposed” (emphasis in the original text).

@@

\section{Informal presentation}
\commentOCf{Je pense changer d’approche : enlever cette section et intégrer les explications informelles directement dans la présentation formelle. À faire…}
\commentYM{Le lecteur n'a aucune idée de ce qu'il doit s'attendre à trouver dans cette section, et de la mesure dans laquelle ça va permettre d'avancer dans la résolution de la problématique : lecteur perdu = pas bien}
\commentYM{par qui ? dans le cadre de l'analyse ou en amont ? et pourquoi sur la seule base de la soundness de l'axiome ? pourquoi pas sur la base de ses implications dans des cas concrets ?}
An empirical theory for a given normative theory, by contrast, makes empirical claims, hence, can be confronted to observations \commentYM{quelle différence avec les implications empiriques d'une théorie normative ?}. What I call here an empirical theory, for simplicity, is not however purely empirical, but can also be considered normative, for two reasons. First, its conclusions inevitably rest on the normative validity of the normative theory for which it is built. Second, it claims to capture preferences that provide an appropriate basis for recommendation. Empirical theories of \ac{DJ} could thus be called empirical normative theories of \ac{DJ}.\commentYM{est-ce la même chose que ce qui est déjà appelé à peu près comme ça dans la littérature ? Je parle par exemple de toutes les théories empiriques de la justice}

\subsection{Normative theories}
Normative theories define an agenda: a set of propositions that we are interested in knowing the deliberated judgment of individuals about ; a set of individuals that the theory claims applies to; a set of arguments;\commentYM{pourquoi est-ce que cela fait partie de la théorie ? c'est plutôt un des éléments du contexte d'une de SES APPLICATIONS à mon sens. L'expression  "normative model" serait selon moi bien plus appropriée si on veut que les détails du cas d'application fassent partie de l'objet. "Theory" a de très fortes connotations de généralité} and a protocol of observation of the judgments an individual adopts towards a proposition after exposure to various arguments \commentYM{pourquoi le protocole d'obs est-il déjà évoqué ici ? j'aurais cru que ce protocole relève de la théorie empirique qui se greffe sur la théorie normative en question. Par ailleurs, je croyais que c'était l'axiome de la théorie normative qui devait être mis à l'épreuve, pas une instantiation sur une propositon. De l'axiome, il n'est plus question. Où est-il ?}.\commentYM{il manque dans la "theorie normative" ce que j'aurais considéré comme le plus important : un modèle de la force de conviction des arguments et de leurs liens avec les propositions. J'aurais dit que l'axiome de la "théorie" se cacherait là-dedans}
